\bf Properties and Bayesian fitting of restricted Boltzmann machines

### 2017-11-30 writing/paper-rbm/docs/paper.tex
submission 2

\title{\bf Properties and Bayes fitting of restricted Boltzmann machines}
\author{Andee Kaplan \\ Department of Statistics, Iowa State University \\  and \\ Daniel Nordman \\ Department of Statistics, Iowa State University \\  and \\ Stephen Vardeman \\ Department of Statistics and Department of Industrial and\\
Manufacturing Systems Engineering, Iowa State University \\ }

@@

\section{Introduction}\label{introduction}

The data mining and machine learning communities have recently shown
great interest in deep learning, specifically in stacked restricted
Boltzmann machines (RBMs) (see R. Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2009deep}{2009}; R. Salakhutdinov
and Hinton \protect\hyperlink{ref-salakhutdinov2012efficient}{2012};
Srivastava, Salakhutdinov, and Hinton
\protect\hyperlink{ref-srivastava2013modeling}{2013}; Le Roux and Bengio
\protect\hyperlink{ref-le2008representational}{2008} for examples). A
RBM is a probabilistic undirected graphical model (for discrete or
continuous random variables) with two layers, one hidden and one
visible, with no conditional dependency within a layer (Smolensky
\protect\hyperlink{ref-smolensky1986information}{1986}). These models
have reportedly been used with success in classification of images
(Larochelle and Bengio
\protect\hyperlink{ref-larochelle2008classification}{2008}; Srivastava
and Salakhutdinov
\protect\hyperlink{ref-srivastava2012multimodal}{2012}). However, the
model properties are largely unexplored in the literature and the
commonly cited fitting methodology remains heuristic-based and relies on
rough approximation (Hinton, Osindero, and Teh
\protect\hyperlink{ref-hinton2006fast}{2006}). In this paper, we provide
steps toward a fuller understanding of the model class and its
properties from the perspective of statistical model theory, and we then
explore the possibility of a rigorous fitting methodology. We find the
RBM model class to be concerning in two fundamental ways.

First, the models can be unsatisfactory as conceptualizations of how
data are generated. That is, recalling Fisher
(\protect\hyperlink{ref-fisher1922mathematical}{1922}), the aim of a
statistical model is to represent data in a compact way. Neyman and Box
further state that a model should ``provide an explanation of the
mechanism underlying the observed phenomena'' (Lehmann
\protect\hyperlink{ref-lehmann1990model}{1990}; G. E. P. Box
\protect\hyperlink{ref-box1967discrimination}{1967}). At issue,
simulation from RBMs can often produce data lacking realistic
variability so that such models may thereby fail to satisfactorily
reflect a true data generation process. Such behavior relates to model
degeneracy (or near degeneracy), which is a statistical concern in that
most data processes being modeled are realistically not degenerate in
their spectrum of potential outcomes. For example, when sampling data
from a nearly degenerate RBM, only a small number of output
possibilities have probability greater than zero, and thus a sample of
images will all be copies of the same one image (or small number of
images). An example of ten 4-pixel images simulated from a nearly
degenerate RBM model is compared to ten 4-pixel images simulated from a
non-degenerate RBM model in Figure \ref{fig:sample-models}. The
degenerate model places almost all probability on one outcome, causing
the image to be generated repeatedly, whereas the non-degenerate model
shows more potentially realistic variation. \par
\begin{figure}
\includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-1} \includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-2} \caption{Ten 4-pixel images simulated from a degenerate model (a) compared to ten 4-pixel images simulated from a non-degenerate model (b). The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more realistic variation.}\label{fig:sample-models}
\end{figure}
In addition to such degeneracy, we find that RBMs can easily exhibit
types of instability, related to how sensitive the model probability
structure can be to small changes in data outcomes. In practice, this
may be seen when a single pixel change in an image results in a wildly
different classification in an image classification problem. Occurrences
of such behavior have recently been documented in RBMs (Li
\protect\hyperlink{ref-li2014biclustering}{2014}), as well as other deep
architectures (Szegedy et al.
\protect\hyperlink{ref-szegedy2013intriguing}{2013}; Nguyen, Yosinski,
and Clune \protect\hyperlink{ref-nguyen2014deep}{2014}). We describe
potential issues of model instability, degeneracy and uninterpretability
for the RBM class, which are related properties, and examine the
presence of these in Section
\ref{explorations-of-model-properties-through-simulation} through
simulations of small, manageable examples.

@@

Notions of weight-decay (penalization) and sparsity (regularization)
have been suggested in the RBM literature as practical remedies for both
over-fitting and poor mixing in the Markov chain during fitting (Hinton
\protect\hyperlink{ref-hinton2010practical}{2010}; Tieleman
\protect\hyperlink{ref-tieleman2008training}{2008}; Cho, Ilin, and Raiko
\protect\hyperlink{ref-cho2012tikhonov}{2012}). Both \(L_1\) and \(L_2\)
type penalties are mentioned to achieve weight-decay, though the
benefits of these particular forms are unknown in different situations.
However, the degree to which regularization and penalization are used by
practitioners is not clear because these concepts are not (by default) a
part of the accepted fitting methodology (Hinton
\protect\hyperlink{ref-hinton2002training}{2002}; Carreira-Perpinan and
Hinton \protect\hyperlink{ref-carreira2005contrastive}{2005}). In this
paper, we attempt to address the concerns with overfitting and poor
mixing in an alternative, and perhaps more transparent manner, via
specification of a Bayesian prior in Section
\ref{bayesian-model-fitting}.

This paper is structured as follows. Section \ref{background} formally
defines the RBM including the joint distribution of hidden and visible
variables and explains the model's connection to deep learning.
Additionally, measures of model impropriety and methods of
quantifying/detecting it are defined. Section
\ref{explorations-of-model-properties-through-simulation} details our
explorations into the model behavior and potential propriety issues with
the RBM class. We examine three Bayesian fitting techniques intended to
avoid model impropriety issues raised in Section \ref{model-fitting} and
conclude with a discussion in Section \ref{discussion}. Appendix
\ref{appendix-rbm} provides proofs for results on RBM parameterizations
and data codings described in Section \ref{data-encoding}.

While applications of the RBM have claimed some successes in
classification problems, it is unclear if the model class allows one to
go beyond fitting to other statistical matters of importance in using
the models, such as quantification of the uncertainty of estimation and
the formulation of predictive distributions. These are closely tied to
the probability properties of the RBM model for explaining data
generation and the current exposition intends to contribute to a better
understanding in this direction.

\section{Background}\label{background}

\subsection{Restricted Boltzmann
machines}\label{restricted-boltzmann-machines}

@@

\subsubsection{Joint distribution}\label{joint-distribution}

@@

\subsubsection{Connection to Deep
Learning}\label{connection-to-deep-learning}

@@

\subsection{Degeneracy, instability, and
uninterpretability}\label{degeneracy-instability-and-uninterpretability}

@@

\subsubsection{Near-degeneracy}\label{near-degeneracy}

@@

A RBM model can be thought to exhibit an analogous form of
\emph{near-degeneracy} when there is a disproportionate amount of
probability placed on a small number of elements in the sample space of
visibles and hiddens, \(\mathcal{C}^{\nv + \nh}\). Using the idea of
Handcock (\protect\hyperlink{ref-handcock2003assessing}{2003}), when the
random vector
\(\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}\)
from (\eqref{eq:t}), appearing in the neg-potential function
\(Q_{\boldsymbol \theta}(\cdot)\), has a mean vector
\(\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}\)
close to the boundary of the convex hull of \(\mathcal{T}\), then the
RBM model can be said to exhibit near-degeneracy at
\(\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}\). Here the
mean of \(\boldsymbol t(\boldsymbol x)\) is
\begin{align*}
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  &= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
\end{align*}
\subsubsection{Instability}\label{instability}

Considering exponential families of distributions, Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) introduced
a concept of model deficiency related to \emph{instability}. Instability
can be roughly thought of as excessive sensitivity in the model, where
small changes in the components of potential data outcomes,
\(\boldsymbol x\), may lead to substantial changes in the probability
mass function \(f_{\boldsymbol \theta}(\boldsymbol x)\). Furthermore,
model instability can be viewed on a spectrum of potential sensitivity
in probability structure, with model degeneracy included as a extreme or
limiting case of instability. To quantify ``instability'' more
rigorously (particularly beyond the definition given by Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011})) it is
useful to consider how RBM models might be expanded to incorporate more
and more visibles. When increasing the size of RBM models, it becomes
necessary to grow the number of model parameters (and in this process
one may also arbitrarily expand the number of hidden variables used). To
this end, let
\(\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1\),
denote an element of a sequence of RBM parameters indexed by the number
\(\nv\) of visibles \((V_1, \dots, V_\nv)\) and define a (scaled)
extremal log-probability ratio of the RBM model at
\(\boldsymbol \theta_{\nv}\) as
\begin{align}
\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) \label{eq:elpr}
\end{align}
where
\(P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{(h_1, \dots, h_\nh) \in \mathcal{C}^\nh}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)\)
is the RBM probability of observing outcome \((v_1, \dots, v_\nv)\) for
the visible variables \((V_1, \dots, V_\nv)\) under parameter vector
\(\boldsymbol \theta_\nv\), after marginalization of hidden variables.

In formulating a RBM model for a potentially large number of visibles
(i.e., as \(\nv \rightarrow \infty\)), we will say that the ratio
(\eqref{eq:elpr}) needs to stay bounded for a sequence of RBM models to be
stable. That is, we make the following convention.
\BeginKnitrBlock{definition}[S-unstable RBM]
\protect\hypertarget{def:unnamed-chunk-1}{}{\label{def:unnamed-chunk-1}
\iffalse (S-unstable RBM) \fi{} }Let
\(\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh + \nh*\nv}, \nv \ge 1\),
be an element of a sequence of RBM parameters where the number of
hiddens, \(\nh \equiv \nh(\nv) \ge 1\), can be an arbitrary function of
the number \(\nv\) of visibles. A RBM model formulation is
\emph{Schweinberger-unstable} or \emph{S-unstable} if
\begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{ELPR}(\boldsymbol \theta_\nv) = \infty.
\end{align*}
\EndKnitrBlock{definition} In other words, a RBM model sequence is
unstable if, given any \(c > 0\), there exists an integer \(n_c > 0\) so
that \(\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c\) for all
\(\nv \ge n_c\). This definition of \emph{S-unstable} is a
generalization or re-interpretation of the ``unstable'' concept of
Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) in that
here RBM models for visibles \((v_1, \dots, v_\nv)\) do not form an
exponential family and the dimensionality of \(\boldsymbol \theta_\nv\)
is not fixed, but rather grows with \(\nv\).

@@

\BeginKnitrBlock{proposition}
\protect\hypertarget{prp:instab1}{}{\label{prp:instab1}}Let \(c > 0\) and
let \(\text{ELPR}(\boldsymbol \theta_\nv)\) be as in (\eqref{eq:elpr}) for
an integer \(\nv \ge 1\). If
\(\frac{1}{\nv}\text{ELPR}(\boldsymbol \theta_\nv) > c\), then
\(\Delta(\boldsymbol \theta_\nv) > c\).
\EndKnitrBlock{proposition}

@@

Additionally, S-unstable RBM model sequences may be formally connected
to the near-degeneracy of Section \ref{near-degeneracy} (in which model
sequences place all probability on a small portion of their sample
spaces). To see this, define an arbitrary modal set of possible outcomes
(i.e.~set of highest probability outcomes) in RBM models with parameters
\(\boldsymbol \theta_\nv, \nv \ge 1\) as \[
M_{\epsilon, \boldsymbol \theta_\nv} \equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v) > (1-\epsilon)\max\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
\] for a given \(0 < \epsilon < 1\). Then S-unstable model sequences are
guaranteed to be degenerate, as the following result shows.

@@

In other words, S-unstable RBM model sequences are guaranteed to stack
up all probability on a specific set of outcomes for visibles, which
could potentially be arbitrarily narrow. Proofs of Propositions
\ref{prp:instab1} and \ref{prp:degen} appear in Kaplan, Nordman, and
Vardeman (\protect\hyperlink{ref-kaplan2017note}{2017}). These findings
also have counterparts in results in Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}), but unlike
results there, we do not limit consideration to exponential family forms
with a fixed number of parameters.

\subsubsection{Uninterpretability}\label{uninterpretability}

@@

\] and

@@

@@

simulation}\label{explorations-of-model-properties-through-simulation}

@@

\subsection{Tiny example}\label{tiny-example}

To illustrate the ideas of model near-degeneracy, instability, and
uninterpretability in a RBM, we consider first the smallest possible
(toy) example that consists of one visible node \(v_1\) and one hidden
node \(h_1\) that are both binary. A schematic of this model can be
found in Figure \ref{fig:toymodel}. Because it seems most common, we
shall begin by employing \(0/1\) encoding of binary variables (both
\(h_1\) and \(v_1\) taking values in \(\mathcal{C} = \{0,1\}\)).
(Eventually we shall argue in Section \ref{data-encoding} that \(-1/1\)
coding has advantages.)
\begin{figure}[ht]
  \centering
  \resizebox{1cm}{!}{\input{../resources/images/tikz/toymodel.tikz}}
  \caption{A small example restricted Boltzmann machine (RBM), with two nodes, one hidden and one visible.}
  \label{fig:toymodel}
\end{figure}
\subsubsection{Impropriety three ways}\label{impropriety-three-ways}

>>>>

\title{\bf Properties and Bayesian fitting of restricted Boltzmann machines}
\author{Andee Kaplan \\ Department of Statistical Science, Duke University \\  and \\ Daniel Nordman \\ Department of Statistics, Iowa State University \\  and \\ Stephen Vardeman \\ Department of Statistics and Department of Industrial and\\
Manufacturing Systems Engineering, Iowa State University \\ }

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The data mining and machine learning communities have recently shown
great interest in deep learning, specifically in stacked restricted
Boltzmann machines (RBMs) (see R. Salakhutdinov and Hinton
\protect\hyperlink{ref-salakhutdinov2009deep}{2009},
\protect\hyperlink{ref-salakhutdinov2012efficient}{2012}; Srivastava,
Salakhutdinov, and Hinton
\protect\hyperlink{ref-srivastava2013modeling}{2013}; Le Roux and Bengio
\protect\hyperlink{ref-le2008representational}{2008} for examples). A
RBM is a probabilistic undirected graphical model (for discrete or
continuous random variables) with two layers, one hidden and one
visible, with no conditional dependency within a layer (Smolensky
\protect\hyperlink{ref-smolensky1986information}{1986}). These models
have reportedly been used with success in classification of images
(Larochelle and Bengio
\protect\hyperlink{ref-larochelle2008classification}{2008}; Srivastava
and Salakhutdinov
\protect\hyperlink{ref-srivastava2012multimodal}{2012}). However, the
model properties are largely unexplored in the literature and the
commonly cited fitting methodology remains heuristic-based and relies on
rough approximation (Hinton, Osindero, and Teh
\protect\hyperlink{ref-hinton2006fast}{2006}). In this paper, we provide
steps toward a fuller understanding of the model class and its
properties from the perspective of statistical model theory, and we then
explore the possibility of a rigorous fitting methodology. We find the
RBM model class to be concerning in two fundamental ways.

First, the models can be unsatisfactory as conceptualizations of how
data are generated. That is, recalling Fisher
(\protect\hyperlink{ref-fisher1922mathematical}{1922}), the aim of a
statistical model is to represent data in a compact way. Neyman and Box
further state that a model should ``provide an explanation of the
mechanism underlying the observed phenomena'' (Lehmann
\protect\hyperlink{ref-lehmann1990model}{1990}; G. E. P. Box
\protect\hyperlink{ref-box1967discrimination}{1967}). At issue,
simulation from RBMs can often produce data lacking realistic
variability so that such models may thereby fail to satisfactorily
reflect a true data generation process. Such behavior relates to model
degeneracy (or near degeneracy), which is a statistical concern in that
most data processes being modeled are realistically not degenerate in
their spectrum of potential outcomes. For example, when sampling data
from a nearly degenerate RBM, only a small number of output
possibilities have probability greater than zero, and thus a sample of
images will all be copies of the same one image (or small number of
images). An example of ten 4-pixel images simulated from a nearly
degenerate RBM model is compared to ten 4-pixel images simulated from a
non-degenerate RBM model in Figure \ref{fig:sample-models}. The
degenerate model places almost all probability on one outcome, causing
the image to be generated repeatedly, whereas the non-degenerate model
shows more potentially realistic variation.

\par
\begin{figure}
\includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-1} \includegraphics[width=1\linewidth]{paper_files/figure-latex/sample-models-2} \caption{Ten 4-pixel images simulated from a degenerate model (a) compared to ten 4-pixel images simulated from a non-degenerate model (b). The degenerate model places almost all probability on one outcome, causing the image to be generated repeatedly, whereas the non-degenerate model shows more realistic variation.}\label{fig:sample-models}
\end{figure}
In addition to such degeneracy, we find that RBMs can easily exhibit
types of instability, related to how sensitive the model probability
structure can be to small changes in data outcomes. In practice, this
may be seen when a single pixel change in an image results in a wildly
different classification in an image classification problem. Occurrences
of such behavior have recently been documented in RBMs (Li
\protect\hyperlink{ref-li2014biclustering}{2014}), as well as other deep
architectures (Szegedy et al.
\protect\hyperlink{ref-szegedy2013intriguing}{2013}; Nguyen, Yosinski,
and Clune \protect\hyperlink{ref-nguyen2014deep}{2014}). We describe
potential issues of model instability, degeneracy and uninterpretability
for the RBM class (which are related properties) and examine the
presence of these in Section
\ref{explorations-of-model-properties-through-simulation} through
simulations of small, manageable examples.

@@

Notions of weight-decay (penalization) and sparsity (regularization)
have been suggested in the RBM literature as practical remedies for both
over-fitting and poor mixing in the Markov chain during fitting (Hinton
\protect\hyperlink{ref-hinton2010practical}{2010}; Tieleman
\protect\hyperlink{ref-tieleman2008training}{2008}; Cho, Ilin, and Raiko
\protect\hyperlink{ref-cho2012tikhonov}{2012}). Both \(L_1\) and \(L_2\)
type penalties are mentioned to achieve weight-decay, though the
benefits of these particular forms are unknown in different situations.
However, the degree to which regularization and penalization are used by
practitioners is not clear because these concepts are not (by default) a
part of the accepted fitting methodology (Hinton
\protect\hyperlink{ref-hinton2002training}{2002}; Carreira-Perpinan and
Hinton \protect\hyperlink{ref-carreira2005contrastive}{2005}). In this
paper, we attempt to address the concerns with overfitting and poor
mixing in an alternative (and perhaps more transparent) manner, via
specification of a Bayesian prior in Section
\ref{bayesian-model-fitting}.

This paper is structured as follows. Section \ref{background} formally
defines the RBM including the joint distribution of hidden and visible
variables and explains the model's connection to deep learning.
Additionally, measures of model impropriety and methods of
quantifying/detecting it are defined. Section
\ref{explorations-of-model-properties-through-simulation} details our
explorations into the model behavior and potential propriety issues with
the RBM class. We examine three Bayesian fitting techniques intended to
avoid model impropriety issues raised in Section \ref{model-fitting} and
conclude with a discussion in Section \ref{discussion}. Supplementary
online material provides proofs for results on RBM parameterizations and
data codings described in Section \ref{data-encoding}.

While applications of the RBM have been claimed to produce some
successes in classification problems, it is unclear if the model class
allows one to go beyond fitting to other statistical matters of
importance in using the models, such as quantification of the
uncertainty of estimation and the formulation of predictive
distributions. These are closely tied to the probability properties of
the RBM model for explaining data generation and the current exposition
intends to contribute to a better understanding in this direction.

\hypertarget{background}{%
\section{Background}\label{background}}

\hypertarget{restricted-boltzmann-machines}{%
\subsection{Restricted Boltzmann
machines}\label{restricted-boltzmann-machines}}

@@

\hypertarget{joint-distribution}{%
\subsubsection{Joint distribution}\label{joint-distribution}}

@@

\hypertarget{connection-to-deep-learning}{%
\subsubsection{Connection to Deep
Learning}\label{connection-to-deep-learning}}

@@

\hypertarget{degeneracy-instability-and-uninterpretability}{%
\subsection{Degeneracy, instability, and
uninterpretability}\label{degeneracy-instability-and-uninterpretability}}

@@

\hypertarget{near-degeneracy}{%
\subsubsection{Near-degeneracy}\label{near-degeneracy}}

@@

A RBM model can be thought to exhibit an analogous form of
\emph{near-degeneracy} when there is a disproportionate amount of
probability placed on a small number of elements in the sample space of
visibles and hiddens, \(\mathcal{C}^{\nv + \nh}\). Using the idea of
Handcock (\protect\hyperlink{ref-handcock2003assessing}{2003}), when the
random vector
\(\boldsymbol t(\boldsymbol x) = (v_1, \dots, v_\nv, h_1, \dots, h_\nh, v_1 h_1, \dots, v_V h_\nh ) \in \mathcal{T} \equiv \{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{C}^{\nh + \nv} \}\)
from (\eqref{eq:t}), appearing in the neg-potential function
\(Q_{\boldsymbol \theta}(\cdot)\), has a mean vector
\(\boldsymbol \mu(\boldsymbol \theta) \in \mathbb{R}^{\nv + \nh + \nv * \nh}\)
close to the boundary of the convex hull of \(\mathcal{T}\), then the
RBM model can be said to exhibit near-degeneracy at
\(\boldsymbol \theta \in \mathbb{R}^{\nv + \nh + \nh*\nv}\). Here the
mean of \(\boldsymbol t(\boldsymbol x)\) is \begin{align*}
\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol X)  &= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x) f_{\boldsymbol \theta} (\boldsymbol x) \right\} \\
&= \sum\limits_{x \in \mathcal{C}^{\nv + \nh}} \left\{ \boldsymbol t(\boldsymbol x)\dfrac{\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}{\sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right)}\right\}.
\end{align*}
\hypertarget{instability}{%
\subsubsection{Instability}\label{instability}}

Considering exponential families of distributions, Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) introduced
a concept of model deficiency related to \emph{instability}. Instability
can be roughly thought of as excessive sensitivity in the model, where
small changes in the components of potential data outcomes,
\(\boldsymbol x\), may lead to substantial changes in the probability
mass function \(f_{\boldsymbol \theta}(\boldsymbol x)\). Furthermore,
model instability can be viewed on a spectrum of potential sensitivity
in probability structure, with model degeneracy included as a extreme or
limiting case of instability. To quantify ``instability'' more
rigorously (particularly beyond the definition given by Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011})) it is
useful to consider how RBM models might be expanded to incorporate more
and more visibles. When increasing the size of RBM models, it becomes
necessary to grow the number of model parameters (and in this process
one may also arbitrarily expand the number of hidden variables used). To
this end, let
\(\boldsymbol \theta_{\nv} \equiv (\theta_{v_1}, \dots, \theta_{v_\nv}, \theta_{h_1}, \dots, \theta_{h_\nh}, \theta_{11}, \dots, \theta_{\nv \nh}), \nv \ge 1\),
denote an element of a sequence of RBM parameters indexed by the number
\(\nv\) of visibles \((V_1, \dots, V_\nv)\) and define a log-ratio of
extremal probabilities (LREP) of the RBM model at
\(\boldsymbol \theta_{\nv}\) as \begin{align}
\frac{1}{\nv} \log \left[\frac{\max\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}{\min\limits_{(v_1, \dots, v_\nv) \in \mathcal{C}^\nv}P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv)}\right] \equiv \frac{1}{\nv} \text{LREP}(\boldsymbol \theta_\nv) \label{eq:elpr}
\end{align} where
\(P_{\boldsymbol \theta_\nv}(v_1, \dots, v_\nv) \propto \sum_{(h_1, \dots, h_\nh) \in \mathcal{C}^\nh}\exp\left(\sum_{i = 1}^\nv \theta_{v_i} v_i + \sum_{j = 1}^\nh \theta_{h_j} h_j + \sum_{i = 1}^\nv \sum_{j = 1}^\nh \theta_{ij} v_i h_j\right)\)
is the RBM probability of observing outcome \((v_1, \dots, v_\nv)\) for
the visible variables \((V_1, \dots, V_\nv)\) under parameter vector
\(\boldsymbol \theta_\nv\), after marginalization of hidden variables.

In formulating a RBM model for a potentially large number of visibles
(i.e., as \(\nv \rightarrow \infty\)), we will say that the ratio
(\eqref{eq:elpr}) needs to stay bounded for a sequence of RBM models to be
stable. That is, we make the following convention.
\BeginKnitrBlock{definition}[S-unstable RBM]
\protect\hypertarget{def:unnamed-chunk-1}{}{\label{def:unnamed-chunk-1}
\iffalse (S-unstable RBM) \fi{} }Let
\(\boldsymbol \theta_\nv \in \mathbb{R}^{\nv + \nh + \nh*\nv}, \nv \ge 1\),
be an element of a sequence of RBM parameters where the number of
hiddens, \(\nh \equiv \nh(\nv) \ge 1\), can be an arbitrary function of
the number \(\nv\) of visibles. A RBM model formulation is
\emph{Schweinberger-unstable} or \emph{S-unstable} if \begin{align*}
\lim\limits_{\nv \rightarrow \infty} \frac{1}{\nv} \text{LREP}(\boldsymbol \theta_\nv) = \infty.
\end{align*}
\EndKnitrBlock{definition}

In other words, a RBM model sequence is unstable if, given any
\(c > 0\), there exists an integer \(n_c > 0\) so that
\(\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c\) for all
\(\nv \ge n_c\). This definition of \emph{S-unstable} is a
generalization or re-interpretation of the ``unstable'' concept of
Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) in that
here RBM models for visibles \((v_1, \dots, v_\nv)\) do not form an
exponential family and the dimensionality of \(\boldsymbol \theta_\nv\)
is not fixed, but rather grows with \(\nv\).

@@

\BeginKnitrBlock{proposition}
\protect\hypertarget{prp:instab1}{}{\label{prp:instab1} }Let \(c > 0\) and
let \(\text{LREP}(\boldsymbol \theta_\nv)\) be as in (\eqref{eq:elpr}) for
an integer \(\nv \ge 1\). If
\(\frac{1}{\nv}\text{LREP}(\boldsymbol \theta_\nv) > c\), then
\(\Delta(\boldsymbol \theta_\nv) > c\).
\EndKnitrBlock{proposition}

@@

Additionally, S-unstable RBM model sequences may be formally connected
to the near-degeneracy of Section \ref{near-degeneracy} (in which model
sequences place all probability on a small portion of their sample
spaces). To see this, define an arbitrary modal set of possible outcomes
(i.e.~set of highest probability outcomes) in RBM models with parameters
\(\boldsymbol \theta_\nv, \nv \ge 1\) as \[
M_{\epsilon, \boldsymbol \theta_\nv} \equiv \left\{\boldsymbol v \in \mathcal{C}^\nv: \log P_{\boldsymbol \theta_\nv}(\boldsymbol v) > (1-\epsilon)\max\log\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) + \epsilon\min\log\limits_{\boldsymbol v^*}P_{\boldsymbol \theta_\nv}(\boldsymbol v^*) \right\}
\] for a given \(0 < \epsilon < 1\). Then S-unstable model sequences are
guaranteed to be degenerate, as the following result shows.

@@

In other words, S-unstable RBM model sequences are guaranteed to stack
up all probability on a specific set of outcomes for visibles, which
could potentially be arbitrarily narrow. Proofs of Propositions
\ref{prp:instab1} and \ref{prp:degen} follow from more general results
in Kaplan, Nordman, and Vardeman
(\protect\hyperlink{ref-kaplan2016propriety}{2016}). These findings also
have counterparts in results in Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}), but unlike
results there, we do not limit consideration to exponential family forms
with a fixed number of parameters.

\hypertarget{uninterpretability}{%
\subsubsection{Uninterpretability}\label{uninterpretability}}

@@

\] and \begin{align*}

@@

\hypertarget{explorations-of-model-properties-through-simulation}{%

@@

simulation}\label{explorations-of-model-properties-through-simulation}}

@@

\hypertarget{tiny-example}{%
\subsection{Tiny example}\label{tiny-example}}

To illustrate the ideas of model near-degeneracy, instability, and
uninterpretability in a RBM, we consider first the smallest possible
(toy) example that consists of one visible node \(v_1\) and one hidden
node \(h_1\) that are both binary. A schematic of this model can be
found in Figure \ref{fig:toymodel}. Because it seems most common, we
shall begin by employing \(0/1\) encoding of binary variables (both
\(h_1\) and \(v_1\) taking values in \(\mathcal{C} = \{0,1\}\)).
(Eventually we shall argue in Section \ref{data-encoding} that \(-1/1\)
coding has advantages.)
\begin{figure}[ht]
  \centering
  \resizebox{1cm}{!}{\input{../resources/images/tikz/toymodel.tikz}}
  \caption{A small example restricted Boltzmann machine (RBM), with two nodes, one hidden and one visible.}
  \label{fig:toymodel}
\end{figure}
\hypertarget{impropriety-three-ways}{%
\subsubsection{Impropriety three ways}\label{impropriety-three-ways}}

For this small model, we are able to investigate the symptoms of model
impropriety, beginning with near-degeneracy. To this end, recall from
Section \ref{near-degeneracy} that one characterization requires
consideration of the convex hull of possible values of statistics
\(\boldsymbol t(\boldsymbol x)\), \[
\mathcal{T} = \{\boldsymbol t(\boldsymbol x): \boldsymbol x = (v_1, h_1) \in \{0,1\}^2\} \equiv \{(v_1, h_1, v_1 h_1): v_1, h_1 \in \{0,1\}\}
\] appearing in the RBM probabilities for this model. As this set is in
three dimensions, we are able to explicitly illustrate the shape of
boundary of the convex hull of \(\mathcal{T}\) and explore the behavior
of the mean vector
\(\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta} \boldsymbol t(\boldsymbol x)\)
as a function of the parameter vector \(\boldsymbol \theta\). Figure
\ref{fig:toyhull} shows the convex hull of our ``statistic space,''
\(\mathcal{T} \subset \{0,1\}^3\), for this toy problem from two
perspectives (enclosed by the unit cube \([0,1]^3\), the convex hull of
\(\{0,1\}^3\)). In this small model, note that the convex hull of
\(\mathcal{T}\) does not fill the unrestricted hull of \(\{0,1\}^3\)
because of the relationship between the elements of
\(\mathcal{T} = \{(v_1, h_1, v_1 h_1 : v_1, h_1 \in \{0, 1\} \}\) (i.e.
\(v_1 h_1 = 1\) only if \(v_1 = h_1 = 1\)).
